What's in a Name? \\ Linear Temporal Logic \\ Literally Represents Time Lines

### 2023-08-28 paper/main.tex
edits to sections 6-7

We prove the correctness of our translation pipeline. The correctness of the translation from LTL formulas to \Buchi automata is provided by SPOT~\cite{Dur22}. Below, we prove the correctness of (i) the translation from \Buchi automata to $\omega$-regular expressions, (ii) the ($\omega$-)regular expression rewrite rules we apply, and lastly, we outline the correctness of our timeline visualizations.
\subsection{Correctness of Regex Translation}
\begin{lemma}
    % reduce_nfa does its input to output behaviour
    For any NFA $G$ and any state $v$ in $G$ that is neither a start state or a final state, \textbf{reduce\_nfa}$(G, v)$ preserves the regular language accepted by $G$
\end{lemma}
\begin{proof}
    Let $G'$ be the graph of $G$ post reduction by the application of \textbf{reduce\_nfa}$(G, v)$. We show that the trace of every path accepted by $G$ is also accepted by $G'$. Suppose a path accepted by $G$ does not pass through $v$, clearly this is true. Else, suppose it passes through $v$, since $v$ is an interior node, $v$ cannot be the first or last in the path. Thus, for every pass through $v$, let $u\neq v$ be the last node passed before entering $v$, and likewise $w$ be the first node after exiting $v$. We show that the regular language of sub-traces from $u$ to $w$ in $G$ (shown in \cref{fig:uvw-dfa}) is identical to that in $G'$.
    \begin{figure}[h!]
        \centering
        \begin{tikzpicture}
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (u) {$u$};
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (v) [right=of u] {$v$};
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (w) [right=of v] {$w$};
	\draw[thick] (u) edge["$r_{uv}$",-latex] (v) (v) edge["$r_{vv}$",-latex,loop above] (v) (v) edge["$r_{vw}$",-latex] (w);
        \end{tikzpicture}
 %       \includegraphics[scale=0.75]{img/uvw_dfa.pdf}
        \caption{A path from $u$ to $w$ in $G$.}
        \label{fig:uvw-dfa}
    \end{figure}
    Suppose there is no loop at $v$, the path from $u$ to $w$ must simply be $u\xrightarrow{r_{uv}}v\xrightarrow{r_{vw}}w$, with the trace $r=r_{uv}r_{vw}$. If there is a loop edge $r_{vv}$, this edge can be traveled any number of times before exiting $v$, thus the regular expression is $r=r_{uv}r_{vv}^*r_{vw}$. This edge $u\xrightarrow{r}w$ is added in $G'$. Thus, for every path accepted by $G$, for every subpath entering and exiting $v$ from some $u$ to $w$, there exists a corresponding subpath from $u$ to $w$ in $G'$, and thus a corresponding path accepted by $G'$.
\end{proof}

@@

\section{Denouement}
Based on our results, we propose the following questions for discussions and future work:
\begin{itemize}
    \item Can there be more simplifications methods as described in Section~\ref{regex-simplify}? The answer is definitely positive. On one hand, more rule-based simplifications can be discovered by finding more patterns in the generated $\omega$-regular expressions. On the other hand, other simplification are possible: for example, if the generated $\omega$-regular expression is $r_1 + r_2$ where $r_1, r_2$ are not syntactically equivalent (or equivalent up to algebraic rules) $\omega$-regular expressions, but they accept the same set of infinite-length strings, then we can simplify this down to just $r_1$. This reduces the problem to equivalence test on $\omega$-regular expressions, which is decidable. In a similar vein, many more simplifications may apply, but the question is how \emph{useful} are they? In the end, we have to balance the efficiency, simplicity, and understandabillity of the visualization.
    %\item Our tool is effective at translating a wide range of LTL formulas. Even with LTL formulas with complex nesting structure are represented by relatively readable timeline visualizations.
    \item Can we define timelines more formally so that one can do this process reversely (i.e., a tool that converts timeline representation to its corresponding \ltl\ formula)? That would be very useful in practice. However, a direct reverse of the algorithm we presented in this paper may not be viable.
\end{itemize}

@@

\item {\bf Can there be back edges in a limited fashion (i.e., to simplify formulas) in timelines?} Most of our formulas have star-height of 1, so if we limit the back-edge representation to only surrounding a single node, that would simplify most formulas. We can then use the expanded notation for stared formulas larger than a single proposition, thus navigating a trade-off between small representation and limiting the visual complication. The criticism of this approach may be that back edges cause confusion between timeline and automata representations, and that with back edges timelines stop being intuitive with respect to the linearity of time.

>>>>

We prove the correctness of our translation pipeline. The correctness of the translation from LTL formulas to \Buchi automata stems from using SPOT~\cite{Dur22}. Below, we prove the correctness of (i) the translation from \Buchi automata to $\omega$-regular expressions, and (ii) the ($\omega$-)regular expression rewrite rules we apply. Lastly, we outline the correctness of our timeline visualizations.
\subsection{Correctness of Regex Translation}
\begin{lemma}
    % reduce_nfa does its input to output behaviour
    For any NFA $G$ and any state $v$ in $G$ that is neither a start state or a final state, \textbf{reduce\_nfa}$(G, v)$ preserves the regular language accepted by $G$
\end{lemma}
\begin{proof}
    Let $G'$ be the graph of $G$ post reduction by the application of \textbf{reduce\_nfa}$(G, v)$. We show that the trace of every path accepted by $G$ is also accepted by $G'$. Suppose a path accepted by $G$ does not pass through $v$, clearly this is true. Else, suppose it passes through $v$, since $v$ is an interior node, $v$ cannot be the first or last in the path. Thus, for every pass through $v$, let $u\neq v$ be the last node passed before entering $v$, and likewise $w$ be the first node after exiting $v$. We show that the regular language of sub-traces from $u$ to $w$ in $G$ (shown in \cref{fig:uvw-dfa}) is identical to that in $G'$.
    \begin{figure}[h!]
        \centering
        \begin{tikzpicture}
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (u) {$u$};
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (v) [right=of u] {$v$};
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (w) [right=of v] {$w$};
	\draw[thick] (u) edge["$r_{uv}$",-latex] (v) (v) edge["$r_{vv}$",-latex,loop above] (v) (v) edge["$r_{vw}$",-latex] (w);
        \end{tikzpicture}
 %       \includegraphics[scale=0.75]{img/uvw_dfa.pdf}
        \caption{A path from $u$ to $w$ in $G$.}
        \label{fig:uvw-dfa}
    \end{figure}
    Suppose there is no loop at $v$, the path from $u$ to $w$ must simply be $u\xrightarrow{r_{uv}}v\xrightarrow{r_{vw}}w$, with the trace $r=r_{uv}r_{vw}$. If there is a loop edge $r_{vv}$, this edge can be traveled any number of times before exiting $v$, thus the regular expression is $r=r_{uv}r_{vv}^*r_{vw}$. This edge $u\xrightarrow{r}w$ is added in $G'$. Thus, for every path accepted by $G$, for every subpath entering and exiting $v$ from some $u$ to $w$, there exists a corresponding subpath from $u$ to $w$ in $G'$, and thus a corresponding path accepted by $G'$.
\end{proof}

@@

\section{Denouement}
Based on our results, we propose the following questions for discussions and future work:
\begin{itemize}
    \item Can there be more simplification methods as described in Section~\ref{regex-simplify}? The answer is definitely positive. For example, we can add more rule-based simplifications by finding more intuitive patterns in the generated $\omega$-regular expressions. Other simplifications are also possible; for example, if the generated $\omega$-regular expression is $r_1 + r_2$ where $r_1, r_2$ are not syntactically equivalent (or equivalent up to algebraic rules) $\omega$-regular expressions, but they accept the same set of infinite-length strings, then we can simplify this down to just $r_1$. This reduces the problem to equivalence testing on $\omega$-regular expressions, which is decidable. In a similar vein, many more simplifications may apply, but the question is how \emph{useful} are they? In the end, we have to balance the efficiency, simplicity, and understandabillity of each output visualization.
    %\item Our tool is effective at translating a wide range of LTL formulas. Even with LTL formulas with complex nesting structure are represented by relatively readable timeline visualizations.
    \item Can we define timelines more formally so that one can do this process reversely (i.e., a tool that converts timeline representation to its corresponding \ltl\ formula)? That would be very useful in practice. However, a direct reverse of the algorithm we presented in this paper may not be viable.
\end{itemize}

@@

\item {\bf Can there be back edges in a limited fashion (i.e., to simplify formulas) in timelines?} Most of our formulas have a star-height of 1, so if we limit the back-edge representation to only surrounding a single node, that would simplify most formulas. We can then use the expanded notation for stared formulas larger than a single proposition, thus navigating the trade-off between having a small representation and limiting the visual complexity. One criticism of this approach may be that back edges cause confusion between timeline and automata representations, and that with back edges timelines stop being intuitive with respect to the linearity of time.
